Entropy stable reduced order modeling of nonlinear conservation laws

### 2019-09-19 docs/ESDG_ROM/main.tex
updating ROM paper

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We introduce projection-based reduced models of nonlinear conservation laws which are globally conservative and inherit a semi-discrete entropy inequality independently of the choice of basis and choice of parameters.  
\end{abstract}
\end{frontmatter}

@@

The construction of stable ROMs for the compressible Navier-Stokes equations is further complicated by their non-trivial structure.  Thus, practitioners often develop structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and can guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

The approach taken in this work differs in the treatment of nonlinear terms.  In \cite{maboudi2018conservative}, equations are rewritten such that the nonlinear terms involve only polynomial (quadratic and cubic) nonlinearities, which can be evaluated exactly using precomputed matrices.  In this work, we approximate nonlinear terms using hyper-reduction techniques based on empirically computed quadrature rules \cite{an2008optimizing, hernandez2017dimensional}.  The hyper-reduction approach incurs approximation error and additional computational cost; however, it is allows for the generalization of the split forms in \cite{maboudi2018conservative} to nonlinearly entropy stable formulations.  We directly construct entropy stable ROMs for nonlinear conservation laws by combining hyper-reduction with a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew} to produce entropy stable reduced order models.

The paper is organized as follows.  Section~\ref{sec:1} introduces entropy inequalities associated with systems of nonlinear conservation laws.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments verifying theoretical results.

\section{Entropy inequalities for nonlinear conservation laws}
\label{sec:1}
Let $\Omega$ denote some domain with boundary $\partial \Omega$.  Nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_i$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \sum_{i=1}^d \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations or inexact quadrature), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

@@

The reduced order models in this work are constructed using full order models based on entropy stable finite volume methods \cite{tadmor1987numerical}.  For simplicity, we illustrate the construction of full and reduced order models on a periodic 1D domain.  We note that, while the reduced order model will be constructed from an entropy stable scheme, the solution snapshots do not need to be generated by entropy stable full order model.  Throughout this work, we will assume that the both full and reduced models yield ``physically relevant'' solutions, such that the entropy is a convex function.  For example, for the compressible Euler and Navier-Stokes equations, we will assume that the thermodynamic variables (density and energy/temperature/pressure) are positive.  Guaranteeing both positivity and accuracy of numerical solutions for general discretizations remains challenging \cite{guermond2016invariant, guermond2019invariant}, and will be explored in future work.

A key ingredient of the full and reduced order models considered in this work are \textit{entropy conservative} finite volume numerical fluxes.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states.  Then, a two-point numerical flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is entropy conservative if it satisfies the following three conditions
\begin{gather}
\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u}), \qquad \text{(consistency)} \label{eq:esflux}\\
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_R), \qquad \text{(symmetry)} \nonumber\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(entropy conservation)}\nonumber,
\end{gather}
Let the domain be decomposed into $K$ cells of size $\Delta x$, and let ${\LRp{\bm{u}_h}_i}$ denote the vector containing mean values of the conservation variables over the $i$th cell.  An entropy conservative finite volume method results from discretizing the integral form of the conservation law as follows
\begin{align}
&\td{{\LRp{\bm{u}_h}_1}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_2},\LRp{\bm{u}_h}_1)-\bm{f}_S({\LRp{\bm{u}_h}_1},{\LRp{\bm{u}_h}_K})}{\Delta x} = \bm{0}, \label{eq:nonlineqs1d}\\
&\td{{\LRp{\bm{u}_h}_i}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_{i+1}},\LRp{\bm{u}_h}_i)-\bm{f}_S({\LRp{\bm{u}_h}_i},{\LRp{\bm{u}_h}_i-1})}{\Delta x} = \bm{0}, \qquad 1 < i < K \nonumber\\
&\td{{\LRp{\bm{u}_h}_K}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_1},\LRp{\bm{u}_h}_K)-\bm{f}_S({\LRp{\bm{u}_h}_K},{\LRp{\bm{u}_h}_K-1})}{\Delta x} = \bm{0},\nonumber
\end{align}
where periodicity is imposed through the equations for ${\LRp{\bm{u}_h}_1}, {\LRp{\bm{u}_h}_K}$.

We first rewrite the system (\ref{eq:nonlineqs1d}) in a matrix form which will be more amenable to model reduction.  We define the skew-symmetric differentiation matrix $\bm{Q}$ and flux matrix $\bm{F}$ such that
\begin{equation}
\bm{Q} = \frac{1}{2}\begin{bmatrix}
0 & 1 & &\ldots & -1\\
-1 & 0 & 1 &&  \\
& -1 & 0 & 1 &  \\
 & & & \ddots &  \\
1 & &\ldots  & -1 & 0
\end{bmatrix}, \qquad \bm{F}_{ij} = \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\label{eq:Qmat}
\end{equation}
Note that $\bm{Q}\bm{1} = \bm{0}$, that the matrix $\bm{F}$ is symmetric (due to symmetry of $\bm{f}_S$), and that the diagonal of $\bm{F}$ is equal to $\bm{f}(\bm{u})$ due to flux consistency.  For a scalar nonlinear conservation law, the matrix-based formulation of (\ref{eq:nonlineqs1d}) is then equivalent to
\begin{align}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = \bm{0}.
\label{eq:es}
\end{align}
where $\circ$ denotes the Hadamard product, $\bm{1}$ denotes the vector of all ones, and each entry of $\bm{u}_h$ corresponds to a point $\bm{x}_i$ in the domain.  In this setting, $\bm{Q}\circ\bm{F}$ extracts and takes linear combinations of nonlinear flux interactions between different nodal values of $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  The discretization of the nonlinear flux term using $\LRp{\bm{Q}\circ \bm{F}}\bm{1}$ is commonly referred to as \textit{flux differencing} \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.\footnote{The factor of 1/2 present in the definition of $\bm{Q}$ is to make (\ref{eq:es}) consistent with the entropy stable SBP literature.  The factor of 2 in (\ref{eq:es}) can also be derived from the chain rule \cite{chen2017entropy, crean2018entropy}.  }

@@

We next show that the full order model satisfies a discrete conservation of entropy.  
\begin{theorem}
\label{thm:ecfom}
Let $\bm{u}_h(t)$ be a solution of (\ref{eq:essys}) for which the entropy $S(\bm{u}_h)$ is convex at each time $t$.  Then, $\bm{u}_h$ satisfies the semi-discrete conservation of entropy 
\[
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = 0.
\]
\end{theorem}
\begin{proof}
The proof can be found in the literature \cite{tadmor1987numerical, tadmor2003entropy, tadmor2016entropy}.  We present an alternative proof which utilizes matrix properties of entropy stable summation by parts (SBP) schemes \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.  We test (\ref{eq:es}) with the vector of entropy variables $\bm{v}_h = \bm{v}(\bm{u}_h)$
\[
\Delta x \bm{v}_h^T\td{\bm{u}_h}{t} + \bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = 0.
\]
Assuming continuity in time and using the definition of the entropy variables in (\ref{eq:nonlineqs}), $\bm{v}_h^T\td{\bm{u}_h}{t}$ simplifies to
\begin{align*}
{\td{S(\bm{u}_h)}{\bm{u}}}^T\td{\bm{u}_h}{t} = \sum_j \LRp{\td{S(\bm{u}_h)}{\bm{u}}}_j^T\td{\LRp{\bm{u}_h}_j}{t} =  \sum_j \td{S(\LRp{\bm{u}_h}_j)}{t}  =  \bm{1}^T\td{S(\bm{u}_h)}{t}.
\end{align*}
Using skew-symmetry of $\bm{Q}$, the flux term $\bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1}$ yields
\begin{align*}
\sum_{ij} \LRp{\bm{v}_h}_i^T \bm{Q}_{ij} \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} = \sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\end{align*}
Rearranging indices and using the symmetry of $\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_L)$ exposes the entropy conservation condition in the sum
\begin{align*}
&\sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} \\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\LRp{\bm{v}_h}_i-\LRp{\bm{v}_h}_j}^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}\\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\psi\LRp{\LRp{\bm{u}_h}_i} - \psi(\LRp{\bm{u}_h}_j)} = \bm{\psi}^T\bm{Q}\bm{1}-\bm{1}^T\bm{Q}\bm{\psi} = 0,
\end{align*}
where we have used that $\bm{Q}=-\bm{Q}^T$ and $\bm{Q}\bm{1} = \bm{0}$ in the final step.

@@

\subsection{Viscosity and entropy dissipation} 
\label{sec:entropydissipation}
Theorem~\ref{thm:ecfom} shows that the formulation (\ref{eq:essys}) preserves a semi-discrete conservation of entropy.  However, in the presence of shock discontinuities, entropy should be dissipated instead of conserved.  To mimic this at the semi-discrete level, we add entropy dissipation to (\ref{eq:es}) through appropriate viscosity terms.  For example, it was shown in \cite{tadmor2006entropy} that a centered approximation of the Navier-Stokes viscosity is entropy dissipative.  In this work, we apply a simple Laplacian artificial viscosity to each component of the solution \cite{upperman2019entropy}
\begin{equation}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} + \epsilon \bm{K}\bm{u}_h= 0, \qquad \bm{K} = \frac{1}{\Delta x^2} \begin{bmatrix}
1 & -1 & & \\
-1 & 2 & \ddots &\\
 & \ddots & \ddots & -1\\
 &  & -1& 1\\
\end{bmatrix}
\label{eq:esvisc}
\end{equation}
where $\epsilon$ is the visosity coefficient.  This choice of dissipation simplifies descriptions of how to treat viscous regularizations of the solution in an entropy stable fashion.  Future work will explore both physically relevant viscosities and more nuanced artificial dissipation mechanisms.

@@

Let $\LRc{{\phi}_j(x)}_{j=1}^{N}$ denote a reduced basis for each component of the solution, which may be generated using principal orthogonal decomposition (POD) or the reduced basis procedure.  Let $\bm{V}$ denote the generalized Vandermonde matrix whose columns contain evaluations of $\phi_j$ at grid points $x_i$

@@

We approximate values of the solution by $\bm{u}_h = \bm{V}\bm{u}_N$, where $\bm{u}_N$ denote ``modal'' coefficients  of the solution in the reduced basis.  Ignoring viscosity terms for now, plugging this expression into (\ref{eq:es}) and enforcing that the residual is orthogonal to all columns of $\bm{V}$ (Galerkin projection) yields a reduced order system

@@

This is not possible due to the fact that the projected entropy variables $\tilde{\bm{v}}$ are no longer mappings of the grid values of the conservative variables $\bm{u}_h = \bm{V}\bm{u}_N$.  To remedy this, we follow approaches taken in \cite{parsani2016entropy, chan2017discretely} and replace the values of ${\bm{u}_h}$ used to evaluate the flux matrix $\bm{F}$ with the \textit{entropy-projected conservative variables} $\tilde{\bm{u}}$

@@

The entropy projected conservative variables are thus mappings of the projected entropy variables.  We note that $\bm{V}\bm{V}^{\dagger}$ can also be interpreted as a discrete approximation of the $L^2$ projection operator onto the span of the reduced basis $\phi_j$. One can then construct a semi-discretely entropy conservative reduced model.

@@

Additionally, if the reduced basis $\bm{V}$ exactly represents $\bm{1}$, solutions of (\ref{eq:esrom}) conserve global averages of the conservative variables.

@@

The conservation of entropy follows from testing with $\bm{v}_N$ and applying (\ref{eq:esromtime}).  The remaining steps are identical to those of Theorem~\ref{thm:ecfom}.  The global conservation results from testing with $\bm{1}$.  Let $\bm{V}\bm{e}=\bm{1}$, where $\bm{e}$ denote coefficients of $\bm{1}$ in the reduced basis matrix.  Then, 

@@

In the absence of hyper-reduction (where $\bm{P}_t = \bm{V}_t^{\dagger}$), the matrix $\bm{Q}_t$ should be an accurate approximation to $\bm{Q}$.  The following lemma quantifies the relationship between accuracy and choice of test space.

@@

>>>>

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We introduce projection-based hyper-reduced models of nonlinear conservation laws which are globally conservative and inherit a semi-discrete entropy inequality independently of the choice of basis and choice of parameters.  
\end{abstract}
\end{frontmatter}

@@

The construction of stable ROMs for the compressible Navier-Stokes equations is further complicated by the non-trivial structure of the equations.  In response, practitioners have developed structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

The approach taken in this work differs from the existing literature in the treatment of nonlinear terms.  In \cite{maboudi2018conservative}, equations are rewritten such that the nonlinear terms involve only polynomial (quadratic and cubic) nonlinearities, which can be evaluated exactly using precomputed matrices.  In this work, we approximate nonlinear terms using hyper-reduction techniques based on empirically computed quadrature rules \cite{an2008optimizing, hernandez2017dimensional}.  The hyper-reduction approach incurs approximation error and additional computational cost; however, it is allows for the generalization of the split forms in \cite{maboudi2018conservative} to nonlinearly entropy stable formulations.  We directly construct entropy stable ROMs for nonlinear conservation laws by combining hyper-reduction with a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew} to produce entropy stable reduced order models.

The paper is organized as follows.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments which verifying the presented theoretical results.

%\section{Entropy inequalities for nonlinear conservation laws}
%\label{sec:1}

@@

We briefly summarize entropy inequalities associated with systems of nonlinear conservation laws.  Let $\Omega$ denote some domain with boundary $\partial \Omega$.  Nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_i$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \sum_{i=1}^d \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations or inexact quadrature), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

The reduced order models in this work are constructed using full order models based on entropy stable finite volume methods \cite{tadmor1987numerical}, which reproduce a discrete version of the continuous entropy inequality (\ref{eq:entropyineq}).  For simplicity, we illustrate the construction of full and reduced order models on a periodic 1D domain.  We note that, while the reduced order model will be constructed from an entropy stable scheme, the solution snapshots do not need to be generated by entropy stable full order model.  Throughout this work, we will assume that the both full and reduced models yield ``physically relevant'' solutions, such that the entropy is a convex function.  For example, for the compressible Euler and Navier-Stokes equations, we will assume that the thermodynamic variables (density and energy/temperature/pressure) are positive.  Guaranteeing both positivity and accuracy of numerical solutions for general discretizations remains challenging \cite{guermond2016invariant, guermond2019invariant}, and will be explored in future work.

A key ingredient of the full and reduced order models considered in this work are \textit{entropy conservative} finite volume numerical fluxes.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states.  Then, a two-point numerical flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is entropy conservative if it satisfies the following three conditions
\begin{gather}
\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u}), \qquad \text{(consistency)} \label{eq:esflux}\\
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_R), \qquad \text{(symmetry)} \nonumber\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(entropy conservation)}\nonumber,
\end{gather}
These fluxes are used to construct entropy conservative and entropy stable finite volume schemes \cite{tadmor1987numerical, chandrashekar2013kinetic, ray2016entropy}.  Let the domain be decomposed into $K$ cells of size $\Delta x$, and let ${\LRp{\bm{u}_h}_i}$ denote the vector containing mean values of the vector of conservation variables over the $i$th cell.  An entropy conservative finite volume method results from discretizing the integral form of the conservation law as follows
\begin{align}
&\td{{\LRp{\bm{u}_h}_1}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_2},\LRp{\bm{u}_h}_1)-\bm{f}_S({\LRp{\bm{u}_h}_1},{\LRp{\bm{u}_h}_K})}{\Delta x} = \bm{0}, \label{eq:nonlineqs1d}\\
&\td{{\LRp{\bm{u}_h}_i}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_{i+1}},\LRp{\bm{u}_h}_i)-\bm{f}_S({\LRp{\bm{u}_h}_i},{\LRp{\bm{u}_h}_i-1})}{\Delta x} = \bm{0}, \qquad 1 < i < K \nonumber\\
&\td{{\LRp{\bm{u}_h}_K}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_1},\LRp{\bm{u}_h}_K)-\bm{f}_S({\LRp{\bm{u}_h}_K},{\LRp{\bm{u}_h}_K-1})}{\Delta x} = \bm{0},\nonumber
\end{align}
where periodicity is imposed through the equations for ${\LRp{\bm{u}_h}_1}, {\LRp{\bm{u}_h}_K}$.

We first rewrite the system (\ref{eq:nonlineqs1d}) in a matrix form which is more amenable to model reduction.  We define the skew-symmetric differentiation matrix $\bm{Q}$ and flux matrix $\bm{F}$ such that
\begin{equation}
\bm{Q} = \frac{1}{2}\begin{bmatrix}
0 & 1 & &\ldots & -1\\
-1 & 0 & 1 &&  \\
& -1 & 0 & 1 &  \\
 & & & \ddots &  \\
1 & &\ldots  & -1 & 0
\end{bmatrix}, \qquad \bm{F}_{ij} = \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\label{eq:Qmat}
\end{equation}
Note that $\bm{Q}\bm{1} = \bm{0}$, that the matrix $\bm{F}$ is symmetric (due to symmetry of $\bm{f}_S$), and that the diagonal of $\bm{F}$ is equal to $\bm{f}(\bm{u})$ due to flux consistency.  For a scalar nonlinear conservation law, the matrix-based formulation of (\ref{eq:nonlineqs1d}) is then equivalent to
\begin{align}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = \bm{0}.
\label{eq:es}
\end{align}
where $\circ$ denotes the Hadamard product, $\bm{1}$ denotes the vector of all ones, and each entry of $\bm{u}_h$ corresponds to a point $\bm{x}_i$ in the domain.  In this setting, $\bm{Q}\circ\bm{F}$ extracts and takes linear combinations of nonlinear flux interactions between different nodal values of $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  The discretization of the nonlinear flux term using $\LRp{\bm{Q}\circ \bm{F}}\bm{1}$ is commonly referred to as \textit{flux differencing} \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}.\footnote{The factor of 1/2 present in the definition of $\bm{Q}$ is to make (\ref{eq:es}) consistent with the entropy stable SBP literature.  The factor of 2 in (\ref{eq:es}) can also be derived from the chain rule \cite{chen2017entropy, crean2018entropy}.  }

@@

We next show that the full order model satisfies a discrete conservation of entropy.  
\begin{theorem}
\label{thm:ecfom}
Let $\bm{u}_h(t)$ be a solution of (\ref{eq:essys}) for which the entropy $S(\bm{u}_h)$ is convex at each time $t$.  Then, $\bm{u}_h$ satisfies the semi-discrete conservation of entropy 
\[
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = 0.
\]
\end{theorem}
\begin{proof}
The proof can be found in the literature \cite{tadmor1987numerical, tadmor2003entropy, tadmor2016entropy}.  Inspired by entropy stable summation by parts (SBP) schemes \cite{carpenter2014entropy, gassner2016split, chen2017entropy, crean2018entropy, chan2017discretely}, we present an alternative proof which relies only on matrix properties of $\bm{Q}$.  We test (\ref{eq:es}) with the vector of entropy variables $\bm{v}_h = \bm{v}(\bm{u}_h)$
\[
\Delta x \bm{v}_h^T\td{\bm{u}_h}{t} + \bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} = 0.
\]
Assuming continuity in time and using the definition of the entropy variables in (\ref{eq:nonlineqs}), $\bm{v}_h^T\td{\bm{u}_h}{t}$ simplifies to
\begin{align*}
{\td{S(\bm{u}_h)}{\bm{u}}}^T\td{\bm{u}_h}{t} = \sum_j \LRp{\td{S(\bm{u}_h)}{\bm{u}}}_j^T\td{\LRp{\bm{u}_h}_j}{t} =  \sum_j \td{S(\LRp{\bm{u}_h}_j)}{t}  =  \bm{1}^T\td{S(\bm{u}_h)}{t}.
\end{align*}
Using skew-symmetry of $\bm{Q}$, the flux term $\bm{v}_h^T2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1}$ yields
\begin{align*}
\sum_{ij} \LRp{\bm{v}_h}_i^T \bm{Q}_{ij} \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} = \sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}.
\end{align*}
Rearranging indices and using the symmetry of $\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_L)$ exposes the entropy conservation condition in the sum
\begin{align*}
&\sum_{ij} \LRp{\bm{Q}_{ij}-\bm{Q}_{ji}} \LRp{\bm{v}_h}_i^T \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j} \\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\LRp{\bm{v}_h}_i-\LRp{\bm{v}_h}_j}^T  \bm{f}_{S}\LRp{\LRp{\bm{u}_h}_i, \LRp{\bm{u}_h}_j}\\
=&\sum_{ij} \bm{Q}_{ij} \LRp{\psi\LRp{\LRp{\bm{u}_h}_i} - \psi(\LRp{\bm{u}_h}_j)} = \bm{\psi}^T\bm{Q}\bm{1}-\bm{1}^T\bm{Q}\bm{\psi} = 0,
\end{align*}
where we have used that $\bm{Q}=-\bm{Q}^T$ and $\bm{Q}\bm{1} = \bm{0}$ in the final step.

@@

\subsection{Viscosity and entropy dissipation} 
\label{sec:entropydissipation}
Theorem~\ref{thm:ecfom} shows that the formulation (\ref{eq:essys}) preserves a semi-discrete conservation of entropy.  However, in the presence of shock discontinuities, entropy should be dissipated instead of conserved.  To mimic this at the semi-discrete level, we add entropy dissipation to (\ref{eq:es}) through appropriate viscosity terms.  For example, it was shown in \cite{tadmor2006entropy} that a centered approximation of the Navier-Stokes viscosity is entropy dissipative.  In this work, we apply a simple Laplacian artificial viscosity to each component of the solution \cite{upperman2019entropy}
\begin{equation}
\Delta x \td{\bm{u}_h}{t} + 2\LRp{{\bm{Q}}\circ \bm{F}}\bm{1} + \epsilon \bm{K}\bm{u}_h= 0, \qquad \bm{K} = \frac{1}{\Delta x^2} \begin{bmatrix}
1 & -1 & & \\
-1 & 2 & \ddots &\\
 & \ddots & \ddots & -1\\
 &  & -1& 1\\
\end{bmatrix}
\label{eq:esvisc}
\end{equation}
where $\epsilon$ is the visosity coefficient.  This choice is intended to simplify the presentation of entropy stable treatments of diffusion terms.  Future work will analyze both physically relevant viscosities and more nuanced artificial dissipation mechanisms.

@@

Let $\LRc{{\phi}_j(x)}_{j=1}^{N}$ denote a reduced basis for each component of the solution, which may be generated using (for example) principal orthogonal decomposition (POD) or the reduced basis procedure.  Let $\bm{V}$ denote the generalized Vandermonde matrix whose columns contain evaluations of $\phi_j$ at grid points $x_i$

@@

We approximate grid values of the solution by $\bm{u}_h = \bm{V}\bm{u}_N$, where $\bm{u}_N$ denote ``modal'' coefficients  of the solution in the reduced basis.  Ignoring viscosity terms for now, plugging this expression into (\ref{eq:es}) and enforcing that the residual is orthogonal to all columns of $\bm{V}$ (Galerkin projection) yields a reduced system

@@

This is not possible due to the fact that the projected entropy variables $\tilde{\bm{v}}$ are no longer mappings of the grid values of the conservative variables $\bm{u}_h = \bm{V}\bm{u}_N$.  To remedy this, we follow approaches taken in \cite{parsani2016entropy, chan2017discretely} and replace the grid values of ${\bm{u}_h}$ used to evaluate the flux matrix $\bm{F}$ with values of the \textit{entropy-projected conservative variables} $\tilde{\bm{u}}$

@@

The entropy projected conservative variables $\tilde{\bm{u}}$ are thus mappings of the projected entropy variables.  We note that $\bm{V}\bm{V}^{\dagger}$ can also be interpreted as a discrete approximation of the $L^2$ projection operator onto the span of the reduced basis $\phi_j$.  One can then construct a semi-discretely entropy conservative reduced model.  

@@

Additionally, if $\bm{1}$ lies within the range of the reduced basis matrix $\bm{V}$ (e.g., $1$ is in the span of the reduced basis functions $\phi_1, \ldots, \phi_N$), solutions of (\ref{eq:esrom}) conserve global averages of the conservative variables.

@@

The conservation of entropy follows from testing with $\bm{v}_N$ and applying (\ref{eq:esromtime}).  The remaining steps are identical to those of Theorem~\ref{thm:ecfom}.  The global conservation results from testing with $\bm{1}$.  If $\bm{1}$ is in the range of $\bm{V}$, then the exist coefficients in the reduced basis matrix $\bm{e}$ such that $\bm{V}\bm{e}=\bm{1}$.  Then, 

@@

In the absence of hyper-reduction (where $\bm{P}_t = \bm{V}_t^{\dagger}$), the matrix $\bm{Q}_t$ should be an accurate approximation to $\bm{Q}$.  The following lemma quantifies the relationship between accuracy in the $2$-norm and choice of test space.  

@@

 In other words, if $\bm{Q}\bm{V}$ is contained within the range of $\bm{V}_t$, the error in approximating $\bm{Q}\bm{f}$ using the compress-and-projected matrix $\LRp{\bm{V}_t\bm{V}_t^{\dagger}}^T\bm{Q}\bm{V}_t\bm{V}_t^{\dagger}$ scales with the best approximation error $\nor{\bm{f}-\bm{V}_t\bm{V}_t^{\dagger}\bm{f}}$.
